\documentclass[onecolumn,apj]{emulateapj}
\usepackage{ctable}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{mathrsfs}
%\usepackage[figuresright]{rotating}
%\usepackage{rotating}
\usepackage{natbib}
%\usepackage{pdflscape}
%\usepackage{lscape}
%\citestyle{aa}

\def\d{\mathrm{d}}
\def\L{\mathscr{L}}
\def\half{\tfrac{1}{2}}

\definecolor{applegreen}{rgb}{0.55, 0.71, 0.0}
\newcommand{\mep}[1]{{\color{applegreen} \textbf{[MEP:  #1]}}}

\begin{document}

\title{Senior Thesis Draft}
\author{Morgan Presley}

\section{General Outline of Thesis}
\begin{itemize}
	\item Introduction \& Background Theory
		\begin{itemize}
			\item Story of Inflation
			\item Motivation (Why we need inflation)
			\item Original failure of Guth's inflation
			\item Slow roll inflation
			\item Problems with current theory (domination of young bubbles; infinite multiverse)
			\item Current status: trying to impose a measure to make bubbles like us more common
			\item But let's ignore the current problems and I'll show that even the simplest models in the current theory must be very fine-tuned to reproduce the current data
		\end{itemize}
	\item How to quantize complexity / fine tuning
		\begin{itemize}
			\item Look at Latham's paper
		\end{itemize}
	\item Examine tuning of inflationary models
		\begin{itemize}
			\item My reproduction of Latham's results
		\end{itemize}
	\item Examine tuning of ekpyrotic models
		\begin{itemize}
			\item My results
		\end{itemize}
	\item Examine tuning of anamorphic models (maybe)
		\begin{itemize}
			\item Have to wait until paper comes out
		\end{itemize}	
	\item Conclusion
\end{itemize}

\section{Introduction}
Inflationary theory was first proposed by Alan Guth as a possible solution to the horizon and flatness problems \citep{Guth1981}. At the time, the standard cosmological model of the early universe was an adiabatically expanding, radiation-dominated universe. This model had two fundamental problems: the horizon problem and the flatness problem. Simply put, the horizon problem is the realization that the universe is homogenous and isotropic on scales far larger than the particle horizon (the farthest distance from which a point could ever receive information, given the age and expansion rate of the universe). 

\section{Derivation of an Inflationary Theory}
\subsection{Inflationary Friedmann equation}
\label{ssec:InflationaryFriedmann}
To see how one could implement an inflationary theory, let us first examine a non-inflationary Friedmann equation:
\begin{equation}
H^2 = \left ( \frac{\dot a(t)}{a(t)} \right ) ^2 = \frac{8 \pi G}{3} \left [ \frac{\rho^0_m}{a^3} + \frac{\rho^0_r}{a^4} + \Lambda + \frac{\sigma^2}{a^6} + \frac{K}{a^2} \right ]
\end{equation}
where $\rho^0_m$ is the energy density of matter in the universe at time $t=0$ (set to be the present time), $\rho^0_r$ is the current energy of radiation, $\Lambda$ is the cosmological constant, $\sigma$ is the cosmic anisotropy term, and $K$ is the curvature of the universe. As we can see by examining the powers of $a$ in each term, the early universe is dominated by the anisotropy term, $\propto a^{-6}$, while the late universe is dominated by the curvature term (and cosmological constant). Therefore, the simplest way to add inflation to the model of the universe is to add an inflationary term to the Friedmann equation: 
\begin{equation}
H^2 = \left ( \frac{\dot a(t)}{a(t)} \right ) ^2 = \frac{8 \pi G}{3} \left [ \frac{\rho^0_m}{a^3} + \frac{\rho^0_r}{a^4} + \Lambda + \frac{\sigma^2}{a^6} + \frac{K}{a^2} + \frac{\rho^0_\phi}{a^{2\epsilon}} \right ]
\end{equation}
where $\phi^0_\phi$ is the energy density of the inflationary field and $\epsilon$ is a positive exponent that must satisfy $\epsilon>3$ for inflation to dominate in the early universe. \mep{I have in my notes that $\epsilon<1$ for inflation and $\epsilon>3$ for contraction, but that doesn't make sense, so I think that's wrong.} 

Now using this form for the inflationary term, let's examine what an inflation-dominated universe would look like. In a $\phi$-dominated universe, the Friedmann equation becomes:
\begin{equation}
H^2 = \left ( \frac{\dot a}{a} \right ) ^2 = \frac{H_0 \Omega_\phi}{a^{2\epsilon}},
\label{eqn:Friedmann_phi}
\end{equation}
where $H_0$ is the Hubble constant and $\Omega_\phi = \tfrac{8\pi G \rho_\phi}{3 H_0}$. Now, solving this equation for $a(t)$ gives
\begin{equation}
a(t) \propto (\epsilon \sqrt{H_0 \Omega_\phi} t)^{1/\epsilon},
\label{eqn:inflation_a(t)}
\end{equation}
and hence 
\begin{equation}
\ddot a(t) \propto \left ( \frac{1-\epsilon}{\epsilon} \right ) (H_0 \Omega_\phi)^{1/2\epsilon} t^{\tfrac{1-2\epsilon}{\epsilon}}. 
\end{equation}
From this, we can see that the condition for an expanding universe is $\epsilon<1$. \mep{Shoot. So my notes must have been right. But how can a small power in the denominator dominate at small values of a? That doesn't make sense!?!? Also, I know for a fact that the radiation term dominates before matter. So to be consistent with that, higher (negative) powers dominate earlier.} Therefore, we want to find an inflationary theory that produces an inflationary term in the Friedmann equation $\propto a^{-2\epsilon}$ so that the universe will be dominated by expansion at early times. In order for inflation to end, $\epsilon$ must grow to become greater than one, so we know that $\epsilon$ is not merely a constant. 

\subsection{Inflationary Field \& Equation of State}

In order to derive a field theory that will produce such an inflationary term, we turn to the equation of state, $w=\rho/p$, where $\rho$ is the energy density and $p$ is the pressure of the field. The equation of state is related to $\epsilon$ by the equation 
\begin{equation}
\epsilon = \tfrac{3}{2}(w+1). 
\label{eqn:eps_w}
\end{equation}
Thus, an inflationary field must have an equation of state $w<-1/3$. Table \ref{tab:field_scenarios} summarizes the values $w$ and $\epsilon$ and the evolution of $a(t)$ for different types of fields. 

\begin{table}[htbp]
   \centering
   \begin{tabular}{@{} lcccc @{}} % Column formatting, @{} suppresses leading/trailing space
      \toprule
      Field Type & $w=\rho/p$ &  $\epsilon=\tfrac{3}{2}(w+1)$ & $a(t)$ & $\ddot a(t)$ \\
      \midrule
      matter & 0 & $3/2$ & $\propto t^{2/3}$ & $>0$ \\
       radiation & $1/3$ & $2$ & $\propto t^{1/2}$ & $<0$ \\
       $\Lambda$ & $-1$ & $0$ & $\propto e^t$ & $>0$ \\
       inflation & $<-1/3$ & $<1$ & $\propto t^{1/\epsilon}$ & $>0$ \\
       contraction & $>1$ & $>3$ & $\propto t^{1/\epsilon}$ & $<0$ \\
      \bottomrule
   \end{tabular}
   \caption{The values $w$ and $\epsilon$ and the evolution of $a(t)$ for different types of fields.}
   \label{tab:field_scenarios}
\end{table}


From Table \ref{tab:field_scenarios}, we saw that an inflationary field must have an equation of state $w<-1/3$. From this, we want to put constrains on the Lagrangian density $\L$ for the inflationary field. Let $\phi$ be an inflationary field. Its Lagrangian density takes the form:
\begin{equation}
\L = -\half ( \partial_\mu \phi)^2 - V(\phi)
\end{equation}
Using the assumptions that $\phi$ is homogenous (so that $\partial_j \phi = 0$) and isotropic (so that we can use the Minkowski metric $\eta_{\mu\nu}$), we find that the stress-energy tensor of the field is
\begin{equation}
T_{\mu\nu} = -\eta_{\mu 0} \dot \phi^2 + \eta_{\mu\nu}(\half\dot \phi^2+V(\phi)).
\end{equation}
From this, we find the equation of state is
\begin{equation}
w = \frac{\half \dot \phi^2 - V(\phi)}{\half \dot \phi^2 + V(\phi)}.
\label{eqn:w}
\end{equation}
Putting in the constraint on $w$ for inflation, we find that an inflationary field must satisfy 
\begin{equation}
\dot \phi^2 < V(\phi).
\end{equation}
\mep{Is this important? It seems it should be, but we never talked about it before...}

Once we have $\L$, we know that the equation of motion for the field is the Klein-Gordon equation:
\begin{equation}
\ddot \phi + 3H \dot \phi + V'(\phi) = 0,
\label{eqn:KG}
\end{equation}
where $H$ is the Hubble parameter, $V(\phi)$ is the inflationary potential, and a dot indicates a time derivative. In our convention, the reduced Planck mass is set to one, i.e. $M_{pl} = (8\pi G)^{-1/2} \equiv 1$. Adding this to the Friedmann equation: 
\begin{equation}
H = \sqrt{\tfrac{1}{3} \left ( \half \dot \phi^2 + V(\phi) \right )}
\label{eqn:Friedmann}
\end{equation}
completely determines the evolution of a universe dominated by an inflationary potential $V(\phi)$. It will be useful later on to re-write the Friedmann equation in terms of $\epsilon$. Putting together equations (\ref{eqn:Friedmann_phi}) and (\ref{eqn:inflation_a(t)}), we find that
\begin{equation}
\epsilon = -\frac{\dot H}{H}
\label{eqn:Friedmann_epsilon}
\end{equation}

\subsection{Calculating Observable Parameters}
Of course, the whole point of creating an inflationary theory is to be able to reproduce observational data. When observations from experiments such as WMAP and Planck observe the early universe, they measure the power spectrum of density fluctuations in the primordial plasma. \mep{More elaboration on where these come from}

Instead of working in units of time, where the time at the end of inflation is difficult to pinpoint, we choose instead to work in terms of the number of e-folds until the end of inflation, defined as 
\begin{equation}
N = -\int_t^{t_{end}} H \d t \text{\ \ \ or\ equivalently\ \ \ } \d N = -H \d t. 
\label{eqn:defN}
\end{equation}
In these units, the end of inflation is defined to occur at $N=0$, and current observations look at a narrow window centered on $N=60$. Note that as time increases, the number of e-folds decreases. 

Figure \ref{fig:density_fluctuations} shows a schematic of the density fluctuations as a function of the wavenumber of the fluctuations ($\log k\sim\tfrac{1}{N}$). If the fluctuations were purely scale invariant, then the curve would be a horizontal line, the amplitude of the fluctuations independent of their wave number. Current observations see only a very limited range near $N=60$, so they focus on measuring (and we focus on predicting) the deviation from scale invariance at that point. This deviation is called the spectral tilt, denoted $n_s(k^*)$, where $k^*$ is the value of $k$ at $N=60$. 



\subsection{Fine-tuning in an Inflationary Theory}
Now that we have the basic mechanics of an inflationary theory, we see that defining an instance of an inflationary theory reduces to defining the inflationary potential $V(\phi)$. 


\section{Eqns for Code}
Let $\phi$ be an inflationary field governed by equations (\ref{eqn:Friedmann}) and (\ref{eqn:KG}). For the purposes of our code, we find it easier to work with $\phi$ not as a function of time but rather as a function of the number of e-folds until the end of inflation, defined in equation (\ref{eqn:defN}). If we let $\phi'$ denote the derivative of $\phi$ with respect to $N$, so that $\dot \phi = -H \phi'$ the Friedmann equation becomes
\begin{equation}
H^2 = \frac{V(\phi)}{3 - \half(\phi')^2},
\end{equation}
and the Klein-Gordon equation becomes 
\begin{equation}
2H^2\phi'' + H^2(\phi')^3 - 6H^2\phi' + 2V'(\phi) = 0.
\label{eqn:KG_N}
\end{equation}

\subsection{Slow Roll Approximation}
In the slow roll approximation, we assume that $\dot \phi$ and $\ddot \phi$ are small. In order for this to be true, the inflationary potential must not be varying rapidly, leading us to the slow roll approximations:
\begin{equation}
\left | \frac{V'(\phi)}{V(\phi)} \right | <<1 \text{\ \ \ and\ \ \ } \left | \frac{V''(\phi)}{V(\phi)} \right | <<1 .
\label{eqn:SlowRollApprox}
\end{equation}
With these approximations, the Friedmann equation becomes
\begin{equation}
3 H^2 \approx_{sr} V(\phi),
\label{eqn:Friedmann_sr}
\end{equation}
and the Klein-Gordon equation becomes
\begin{equation}
3 H \dot \phi \approx_{sr} -V'(\phi)
\label{eqn:KG_sr}
\end{equation}
We use these approximations to find the initial conditions for integrating equation (\ref{eqn:KG_N}): $\phi(N=0)$ and $\dot \phi(N=0)$. To do this, we use the fact that, at the end of inflation (i.e. at $N=0$), $\epsilon = 1$. Recall from Section \ref{ssec:InflationaryFriedmann} that the condition for accelerated expansion is $\epsilon<1$. Therefore, as the universe transitions out of the inflationary epoch and into a period of decelerating expansion (where $\epsilon>1$), there must be a point where $\epsilon=1$, marking the end of inflation. Putting together equations (\ref{eqn:eps_w}) and (\ref{eqn:w}), we find that 
\begin{equation}
\epsilon = \frac{3}{2} \left ( \frac{\dot \phi^2}{\half \dot \phi^2 + V(\phi)} \right ) \approx_{sr} \half \left (\frac{V'(\phi)}{V(\phi)} \right ) ^2 
\end{equation}
where we have used the slow roll approximations $\half \dot \phi^2 << V(\phi)$ and equation (\ref{eqn:KG_sr}). Setting $\epsilon=1$ gives us an equation we can solve for $\phi(N=0)$. Once we have $\phi(N=0)$, we can easily us equation (\ref{eqn:KG_sr}) to find $\dot \phi(N=0)$. 

\bibliographystyle{apj}
\bibliography{thesis}{}

\end{document}